Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-27 manuscript.Rmd
add updated descriptives table to repo
revisions to discussion, conclusion

In this approach, there are two metrics that each reflect a combination of relevant theoretical processes. The first metric, interpersonal differences at the start of the survey, reflect the accumulated experiences of people prior to entering the panel survey that shape how they answer questions, what Lersch calls "early imprinting." While commonly associated with experiences during an early formative period that result in settled dispositions, this variation could also reflect experiences that happen at any stage in life as long as they predate the panel and consistently impact subsequent responses. For instance, for those people entering the panel post-retirement, this “imprinting” might reflect this pivotal life transition. Consequently, this metric also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions. [TK Note: I feel that this paragraph puts too much emphasis on pre-panel processes. It might be early imprinting or broad life-course, but why emphasize it this much? KK NOTE: Because it could be either, I don't want to treat it like it's just "imprinting."]
[KK NOTE: You reversed what was being said here. You say "early imprinting" denotes intrapersonal change (process denotes number). We want to say the opposite -- what I had before -- number denotes process: "interpersonal differences" denotes "early imprinting." I changed it back. I also took out reference to "transformative" experiences because I don't think we're necessarily talking about those specifically, just any experience that changes you. Because we don't say a specific theoretical process ("socialization", etc.) I want to keep the language of each denoting combinations of processes. "Persistent change" isn't a theoretical process. It's a bunch of them.]

@@

Second, we expect the variance explained by intrapersonal to be modest compared to interpersonal differences at the start of the survey. This expectation should be uncontroversial, regardless of the theoretical process assumed to be most relevant for a particular question. Panel surveys do not encompass individuals’ entire lifetimes. Thus, even if intrapersonal change during adulthood is the major driver of cultural differences, the change detected in a panel covering 20 years probably will not fully account for the entire variance of interpersonal differences.

@@

We use data from 7 nationally representative panel surveys from Australia, Germany, Great Britain, Switzerland, and the United States (summarized in Table 1), combining all the data files used in previous work [@kiley2020; @lersch2023].[^data] These studies cover a long period of time (the range of surveys spans from 1968 to 2021), with 609 personal culture items that capture attitudes, beliefs, values, self-assessments, self-descriptions, and behaviors [@alwin2007]. We restricted the sample such that individuals between the ages of 18 and 79 are included without further elimination, and in all surveys, we used all possible cases for which respondents provided responses. In the end, the analyses that follow rely on a cross-cultural sample with a cross-domain set of items to capture the breadth of individual personal culture. Appendix A documents the list of all variables used in the upcoming analyses.

@@

$$
y_{i, t} = \beta_0 + u_{0, i} + \beta_1 x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
$$

@@

As a first step in our analysis, we fit LCAMs to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of systematic variance that correspond to the two collections of theoretical processes outlined above: interpersonal differences at baseline and intrapersonal change over time. First, to measure the importance of interpersonal differences, we calculate the sum of the squared residuals derived from measuring each respondent at the midpoint of their age trajectory.[^midpoint] We then divide this value by the total sum of squares and subtract it from from 1, producing a value we denote as V(D), or the proportion of variance attributable to interpersonal differences. This number is equivalent to an $R^2$ value generated by using random intercepts for all respondents. Second, we calculate the sum of squared residuals generated by additionally accounting for the age trajectory. We divide this number by the total sum of squares and subtract it from 1. We then subtract $V(D)$ from this value. This gives us a value we denote as $V(C)$, the additional proportion of variance explained by intrapersonal change.

@@

We can think of this number, $\tau$ as akin to an Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher the number is, the more variance in responses is attributable to intrapersonal change during the panel. The lower the number, the less intrapersonal change seems to play a large role in explaining cultural differences during the panel. Moving forward, we use $\tau$ as a placeholder for *systematic variance attributable to intrapersonal change*.

Because this quotient summarizes variance and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have very different levels of non-systematic variance, giving us a baseline sense of the relative prevalence of these two processes for each question (something not achievable under the tournament of models approach), and allowing us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

@@

[TK Note: I guess we need to say something here about the college stuff? KK Note: I didn't feel like that was necessary. It feels so distinct later on that I think it's OK to kind of "start over" there. But others should weigh in.].

@@

Figure 1 plots the proportion of systematic variance attributable to interpersonal differences and intrapersonal change for each of the 609 questions analyzed, plotted separately by panel. To broadly summarize the results, all panels show a similar range of the proportion of systematic variance attributable to intrapersonal change. Across panels, the median proportion of $\tau$ is 0.094 (mean of 0.098), with an interquartile range of 0.053 to 0.139. All panels include questions where $\tau$ is essentially 0 and all panels include questions with $\tau$ values greater than 0.20.

@@

To the extent that there are differences across the panels, the PSID has the highest $\tau$ values with mean .120 and median .135. While we cannot disentangle features of the sample from features of the questions asked to each sample, the specific samples for many PSID questions have lower average ages than those from other panels. To the extent that younger respondents might be more likely to make durable changes of opinion, this higher explanatory power of intrapersonal change might reflect the distinct age profile of respondents in this sample. At the other end, the GSS has the lowest range of $\tau$ with mean .073 and median .069. This potentially reflects the fact that the GSS observes people for a shorter duration, on average, than the other panels. If, consistent with life course adaption theories, people are more likely to make significant cultural changes the longer we observe them, then duration likely affects the range of $\tau$ (a point we explore more below). However, the GSS results are still highly consistent with results from the other panels.

@@

Appendix A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, as well as the proportion of residual variance, for each question analyzed. In general, interpersonal differences are almost always the largest component of the total variance and tend to explain between 55 and 70 percent of total variance, while intrapersonal change is always the smallest, typically explaining between 3 and 8 percent of total variance. Residual variance tends to account for between 22 and 37 percent of variance (though on several questions residual variance is greater than 50 percent).

@@

Figure 2 plots the results from an ordinary least squares regression of $\tau$ as a function of question, panel, and sample features. These models also include fixed effects for panels (not shown), so coefficients reflect the association between a particular sample or question characteristic within a panel.[KK NOTE: CAN WE SUMMARIZE THESE PANEL FIXED EFFECTS AT ALL? AE: WE COULD SHOW THAT THEY DON'T CAPTURE TOPIC EFFECTS BEYOND OUR CLASSIFICATION IF WE INCLUDE THOSE / THAT THEY POSSIBLY CAPTURE UNMEASURED COUNTRY/CONTEXT DIFFERENCES OR DIFFERENCES IN SURVEY METHODOLOGY?]

@@

This reinforces the explanation for why the GSS displays the lowest average proportion of systematic variance attributable to intrapersonal change: its shorter duration. [KK NOTE: Does this inform any other panel differences? Does the PSID, BHPS or UKHLS have longer durations on average?]

@@

While these broad results help speak to previous debates, they mostly re-frame previous findings without extending the theoretical debate. To better illustrate the value of our approach for addressing different theoretical questions, we turn to a specific empirical example: the relative importance of intrapersonal change and interpersonal difference for explaining cultural differences by education levels.

Previous work has established a positive relationship between education and attitude stability, especially on issues related to American politics. This stability is often attributed to education facilitating "chronic information" -- a general understanding of the domain of American politics, including the positions held by major parties and political figures and how issues relate to one another at a logical or socio-logical level [@zaller1992; @alvarez2002; @boutyline2017]. Because they have more knowledge of American politics, people with a college education are better able to consistently connect the considerations in their cognition with the answer choices they are presented with in a survey.

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences than the rest of the population [@zaller1992; @alwin2007]. However, this focus on the difference in the non-systematic (residual) component of variance across groups obscures the fact that the relative importance of different components of systematic variance might differ across these two groups as well.

It could be the case that education, chronic information, and attention to the political sphere makes intrapersonal change more important in explaining differences among people with higher levels of education. For example, these people might be more likely to make systematic changes in response to the emergence of new information, new issues, or political realignments, while those without chronic information display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. College could serve as a formative experience that solidifies personal culture for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing personal culture, as these experiences potentially provide information that college-educated peers received earlier.

To test these competing propositions, we use the models presented above to calculate the total systematic variance for people with at least a bachelor's degree and people with less than a bachelor's degree at wave 1[^wave] of the General Social Survey's panels.[^missing] We focus on the GSS because it contains the largest number of questions tapping general political dispositions, which is the domain where education has proven particularly relevant for understanding attitude stability. The GSS also covers a particularly turbulent window of American politics from 2006 to 2014. This window covers the start of the Great Recession, debates about federal intervention in and regulation of Wall Street, the election of Barack Obama as the first black president in U.S. history, debates about the role of the federal government in the health care sector, the emergence of the Tea Party, and political realignment and clarification on the issue of gay marriage, among other topics.

Figure 3 plots the distribution of differences in $\tau$ values between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel for all 183 items int he GSS. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

@@

There is a clear pattern in Figure 3: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (i.e., they have negative values). While most of these differences are small in absolute terms (less than 2 percentage points), several are greater than 5 percentage points. Given the distribution observed in Figure 1 showing that the systematic variance attributable to intrapersonal change averages around 0.09, a 5 percentage point difference between groups is quite substantial.

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to blacks, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether blacks should be given preferences in hiring, measured on a five-point scale. Estimates of $\tau$ for these eight questions, for both education groups, are presented in Figure 4.

@@

On all eight questions presented in Figure 4, intrapersonal change explains less systematic variance for people with a college degree than for people without a college degree. There are also large differences in $\tau$ values across questions for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show larger values of $\tau$ than the other questions, as well as a large difference between people with a college degree and those without. Compared to the other questions explored here, change as adults plays a much larger role in explaining why people report the partisan identification and political ideology they do. And this is particularly true for respondents who do not have a college degree. The $\tau$ value for ideological identification among non-college educated respondents is almost four times that of college-educated respondents.

We note here that this meaningful difference in $\tau$ values across education groups and across questions would not be detectable using previous methods. For partisan identification and political ideology, both college-educated respondents and people with less than college degree would likely favor the AUM or LCAM over the SDM because they show evidence of some members of the population making intrapersonal change. In other words, using the tournament of models approach obscures the fact that the relative importance of intrapersonal change in explaining differences on these two items is different across these two groups, that intrapersonal change explains more systematic variance for ideological identification than partisan identification for people without a college degree, and that there appears to be more durable change on questions of affirmative action than on questions of government aid to black Americans.

It is important to note here that these differences should not be interpreted as causal effects of attending college on the importance of social experiences for cultural change. These results are potentially confounded by age, social class, race, gender, and other factors that explain selection into higher education. We simply use these estimates as a starting point for a broader discussion.

@@

This paper set out intervene in the debate on whether people change their personal culture -- their attitudes, beliefs, values, practices, and dispositions -- as they move through their adult life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to call victory for one or the other possible answer to this question, it seeks to refocus the debate to the theoretically more productive question of _how_ initial differences between and durable changes within people collectively explain differences in the personal culture they exhibit. To enable exploring this and related questions, it proposes a new approach that quantifies the amount of observed systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change.

Applying our proposed measure, $\tau$, to all panel datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of durable, intrapersonal change over time, with some showing notably high intrapersonal change compared to interpersonal differences. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

[TK Note: from now on I didn't change anything to $\tau$, but not sure if this is a correct way?].

Intrapersonal change is often substantially less pronounced than interpersonal differences, accounting for less than 10 percent on average across all datasets. On numerous questions, like those on civil liberties, abortion, generalized trust, and civic duty, the systematic variance attributable to intrapersonal change is essentially zero. For these questions, it seems there is not enough cultural change during adulthood to warrant attributing differences in personal culture mainly to experiences and social transitions; instead, the primary source of the observed differences appears to stem from early-life experiences [OR "EXPERIENCES PREDATING ADULT LIFE"].

More than adjudicating past positions in this debate, the approach presented in this paper enables scholars to ask when change or stability in personal culture predominates within a given population and examine their role in explaining the cultural variation within it. To exemplify this, we detailed how the proportion in cultural differences that intrapersonal change can explain not only varies substantially by survey item but also by individuals' characteristics such as their education. Intrapersonal change plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, it is more important in explaining cultural variation among non-college graduates compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, this suggests that experiencing college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions less influential for college graduates relative to non-college graduates.

@@

So far, sociologists interested in understanding cultural differences have largely asked about the presence and significance of cultural change in adults [e.g., @kiley2020; @vaisey2021; @lersch2023]. Yet doing so, has unnecessarily limited and might even set the wrong focus. In any population, some degree of cultural change in adults is inevitable. Likewise, no viable theoretical perspective would expect just one process -- stable differences or internal change -- to shape their personal culture. Our results reinforce this point, showing the relevance of both interpersonal differences and intrapersonal change in understanding the differences in personal culture among adults across all items analyzed. [TRIED TO BETTER TUCK IT IN, BUT COULD LET GO OF THE LAST SENTENCE]

This and other recent findings [CITES] suggest that it is theoretically more productive to ask about the relative importance of these two processes in understanding cultural variation. Drawing a unified conclusion from survey items measuring various cultural forms on different scales and across different time frames is challenging. However, the general pattern suggests that, for most items, intrapersonal change in adulthood is not the primary reason for cultural differences among people.

To clarify, our findings do not imply that adults remain static or that their changes are inconsequential. Even minor shifts in personal culture, such as a 2% change in support for gay marriage, can have significant ramifications. While the majority may remain consistent in their views, understanding the underlying mechanisms of even such minor shifts remains a crucial task for the sociology of culture.

@@

This suggestion aligns with a range of recent causal inference work suggesting ["ADVOCATING" TO AVOID REPETITION] that selection effects, rather than treatment effects, predominantly account for personal cultural differences among individuals in varied social roles and positions [@campbell2016; @wodtke2018]. While some individuals clearly change as they transition into new roles or environments, this change seems, on average, to be insufficient in magnitude and duration to possibly explain the pronounced differences among people in diverse roles. [I'VE TRIED TO FIND A WAY TO GET BOTH IMPACT AND DURATION IN THERE; COULDN'T FIND A BETTER SOLUTION THOUGH--IDEAS? ALSO WE NEED TO BE CAREFUL WITH THIS SENTENCE, IT SEEMS WE ARE OVERSTATING HERE; LET'S FIND A WAY TO RELATIVIZE THE CLAIM HERE?]. When observing differences in personal culture across social roles, such as parenthood, education, or professional authority [@longest2013], or across occupations [@weeden2005], scholar are well advised to expect selection as their primary cause, rather than treatment [ALTERNATIVE: "OUR INITIAL PRESUMPTION SHOULD LEAN TOWARDS..."].

Our analysis does not provide an answer as to why intrapersonal change seems to have limited impact on understanding cultural differences. The situations that promote durable change in personal culture might simply be rare. Alternatively, it is possible that adults do encounter opportunities, necessities, and incentives for change, but their ability or willingness to change decreases. All the more it is important to research when and how social situations can provoke durable change in adults.

>>>>

In this approach, there are two metrics that each reflect a combination of relevant theoretical processes. The first metric, interpersonal differences at the start of the survey, reflects the accumulated experiences of people prior to entering the panel survey that shape how they answer questions, what Lersch calls "early imprinting." While commonly associated with experiences during an early formative period that result in settled dispositions, this variation could also reflect experiences that happen at any stage in life as long as they predate the panel and consistently impact subsequent responses. For instance, for those people entering the panel post-retirement, this “imprinting” might reflect this pivotal life transition. Consequently, this metric also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions. [TK Note: I feel that this paragraph puts too much emphasis on pre-panel processes. It might be early imprinting or broad life-course, but why emphasize it this much? KK NOTE: Because it could be either, I don't want to treat it like it's just "imprinting."]
[KK NOTE: You reversed what was being said here. You say "early imprinting" denotes intrapersonal change (process denotes number). We want to say the opposite -- what I had before -- number denotes process: "interpersonal differences" denotes "early imprinting." I changed it back. I also took out reference to "transformative" experiences because I don't think we're necessarily talking about those specifically, just any experience that changes you. Because we don't say a specific theoretical process ("socialization", etc.) I want to keep the language of each denoting combinations of processes. "Persistent change" isn't a theoretical process. It's a bunch of them.]

@@

Second, we expect the variance explained by intrapersonal change to be modest compared to interpersonal differences at the start of the survey. This expectation should be uncontroversial, regardless of the theoretical process assumed to be most relevant for a particular question. Panel surveys do not encompass individuals’ entire lifetimes. Thus, even if intrapersonal change during adulthood is the major driver of cultural differences, the change detected in a panel covering 20 years probably will not fully account for the entire variance of interpersonal differences.

@@

We use data from 7 nationally representative panel surveys from Australia, Germany, Great Britain, Switzerland, and the United States (summarized in Table 1), combining all the data files used in previous work [@kiley2020; @lersch2023].[^data] These studies cover a long period of time (the range of surveys spans from 1968 to 2021), with 609 personal culture items that capture attitudes, beliefs, values, self-assessments, self-descriptions, and behaviors [@alwin2007]. We restricted the sample such that individuals between the ages of 18 and 79 are included without further elimination, and in all surveys, we used all possible cases for which respondents provided responses. In the end, the analyses that follow rely on a cross-cultural sample with a cross-domain set of items to capture a broad range of individual personal culture. Appendix A documents the list of all variables used in the upcoming analyses.

@@

$$
y_{i, t} = \beta_0 + u_{0, i} + \beta_{1} x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
$$
$$
u_{0,i} \sim \mathcal{N}(0,\sigma_{\mu_0}^2)
$$

$$
u_{1,i} \sim \mathcal{N}(0,\sigma_{u_1}^2)

$$
[KKNOTE: Is that right? I forget how to do that exactly.]

@@

As a first step in our analysis, we fit a LCAM to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of systematic variance that correspond to the two collections of theoretical processes outlined above: interpersonal differences at baseline and intrapersonal change over time. First, to measure the importance of interpersonal differences, we calculate the sum of the squared residuals derived from measuring each respondent at the midpoint of their age trajectory.[^midpoint] We then divide this value by the total sum of squares and subtract it from from 1, producing a value we denote as V(D), or the proportion of variance attributable to interpersonal differences. This number is equivalent to an $R^2$ value generated by using random intercepts for all respondents. Second, we calculate the sum of squared residuals generated by additionally accounting for the age trajectory. We divide this number by the total sum of squares and subtract it from 1. We then subtract $V(D)$ from this value. This gives us a value we denote as $V(C)$, the additional proportion of variance explained by intrapersonal change.

@@

We can think of this number, $\tau$ as akin to an Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher this number is, the more systematic variance in responses is attributable to intrapersonal change during the panel. The lower the number, the less intrapersonal change seems to play a large role in explaining systematic cultural differences during the panel. Moving forward, we use $\tau$ as a placeholder for *systematic variance attributable to intrapersonal change*.

Because this quotient summarizes variance and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have different levels of non-systematic variance, giving us a sense of the relative prevalence of these two processes for each question (something not achievable under the tournament of models approach), and allowing us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

@@

[TK Note: I guess we need to say something here about the college stuff? KK Note: I didn't feel like that was necessary. It feels so distinct later on that I think it's OK to kind of "start over" there. I'm worried that people will forget the plan by the time they get there, and we'll have to reiterate everything anyway. But others should weigh in.].

@@

Figure 1 plots the proportion of systematic variance attributable to interpersonal differences and intrapersonal change for each of the 609 questions analyzed, plotted separately by panel. To broadly summarize the results, all panels show a similar range of the proportion of systematic variance attributable to intrapersonal change. Across panels, the median proportion of $\tau$ is 0.094 (mean of 0.098), with an interquartile range of 0.053 to 0.139. All panels include questions where $\tau$ is essentially 0, and all panels include questions with $\tau$ values greater than 0.20. The maximum $\tau$ value is 0.26.

@@

To the extent that there are differences across the panels, the PSID has the highest $\tau$ values with mean .120 and median .135. While we cannot disentangle features of the sample from features of the questions asked to each sample, the specific samples for many PSID questions have lower average ages than those from other panels. To the extent that younger respondents might be more likely to make durable changes of opinion, this higher explanatory power of intrapersonal change might reflect the distinct age profile of respondents in this sample. At the other end, the GSS has the lowest range of $\tau$ with mean .073 and median .069. This potentially reflects the fact that the GSS observes people for a shorter duration, on average, than the other panels. If, consistent with life course adaption theories, people are more likely to make significant cultural changes the longer we observe them, then duration likely affects the range of $\tau$ (a point we explore more below). However, the GSS results are still consistent with results from the other panels.

@@

Appendix A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, the estimated values of $\tau$, and the proportion of residual variance for each question analyzed. In general, interpersonal differences are almost always the largest component of the total variance and tend to explain between 55 and 70 percent of total variance, while intrapersonal change is always the smallest, typically explaining between 3 and 8 percent of total variance. Residual variance tends to account for between 22 and 37 percent of variance, though on several questions residual variance is greater than 50 percent.

@@

Figure 2 plots the results from an ordinary least squares regression of $\tau$ as a function of question, panel, and sample features. These models also include fixed effects for panels, so coefficients reflect the association between a particular sample or question characteristic within a panel.[KK NOTE: CAN WE SUMMARIZE THESE PANEL FIXED EFFECTS AT ALL? AE: WE COULD SHOW THAT THEY DON'T CAPTURE TOPIC EFFECTS BEYOND OUR CLASSIFICATION IF WE INCLUDE THOSE / THAT THEY POSSIBLY CAPTURE UNMEASURED COUNTRY/CONTEXT DIFFERENCES OR DIFFERENCES IN SURVEY METHODOLOGY?]

@@

This reinforces the explanation for why the GSS displays the lowest average proportion of systematic variance attributable to intrapersonal change: its shorter duration.

While we did not generate expectations for how question content relates to $\tau$, some of the associations between question structure, panel duration, and panels themselves might be driven by differences in the topics addressed by each panel. To address this, we coded each question as falling into one of nine different content categories and included these as indicator variables in the meta analysis. Appendix C presents the estimates for these content indicators. They show some associations with $\tau$, with questions about subjective SES; social life, social cohesion, and trust; environment and climate; and health and morale showing larger $\tau$ values than questions about religion and spirituality; politics, government and the economy; and gender and family. [KKNOTE: I FEEL LIKE WE WANTED TO SAY SOMETHING ELSE HERE BUT I CANT REMEMBER WHAT]

@@

These results mostly re-frame and align previous findings, but the value of our approach lies in its ability to extend the debate to a broader set of theoretical questions. To show this potential, we turn to a specific empirical example: the relative importance of intrapersonal change and interpersonal difference for explaining cultural variation by education levels.

Previous work has established a positive relationship between education and attitude stability, especially on issues related to American politics. This stability is often attributed to education facilitating "chronic information" -- a general understanding of and attention to the domain of American politics, including the positions held by major parties and political figures and how issues relate to one another at a logical or socio-logical level [@zaller1992; @alvarez2002; @boutyline2017]. These perspectives argue that because college graduates have more knowledge of American politics, they are better able to consistently connect the considerations in their cognition with the answer choices they are presented with in a survey.

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences than the rest of the population [@zaller1992; @alwin2007]. But focus on the difference in the non-systematic or residual component of variance across groups obscures the fact that the relative importance of different components of systematic variance, intrapersonal change and interpersonal differences, might also differ across these groups.

There are theoretical reasons to believe that education might be associated with either more or less systematic variance attributable to intrapersonal change. On one hand, because college graduates are more connected to mainstream discourse and elite signals, they might be more likely to make durable changes in response to the emergence of new information, new issues, or political realignments, while those without chronic information display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. In this perspective, college could serve as a formative experience that solidifies some dimensions of personal culture for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing personal culture, as these experiences potentially provide information that college-educated peers received earlier.

To test these competing propositions, we use the models presented above to separately calculate $\tau$ values for people with at least a bachelor's degree and people with less than a bachelor's degree at wave 1[^wave] of the General Social Survey's panels.[^missing] We focus on the GSS because it contains the largest number of questions tapping general political dispositions, which is the domain where education has proven particularly relevant for understanding attitude stability. The GSS also covers a turbulent window of American politics from 2006 to 2014. This window covers the start of the Great Recession, debates about federal intervention in and regulation of Wall Street, the election of Barack Obama as the first black president in U.S. history, debates about the role of the federal government in the health care sector, the emergence of the Tea Party, and political realignment and clarification on the issue of gay marriage, among other topics.

Figure 3 plots the distribution of differences in $\tau$ values between people with at least a bachelor's degree and people with less than a bachelor's degree at wave 1 of the panel for all 183 GSS items we explored. Values greater than 0 indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

@@

There is a clear pattern in Figure 3: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (i.e., the difference is negative). While most of these differences are small in absolute terms, less than 2 percentage points, several are greater than 5 percentage points. Given the distribution observed in Figure 1 showing that the systematic variance attributable to intrapersonal change averages around 0.09, a 5 percentage point difference between groups is quite substantial.

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to black people, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether black people should be given preferences in hiring, measured on a five-point scale. Estimates of $\tau$ for these eight questions, for both education groups, are presented in Figure 4.

@@

On all eight questions presented in Figure 4, $\tau$ values are greater for people without a college degree than for people with a college degree, meaning intrapersonal change explains less systematic variance for people with a college degree than for people without one. There are also large differences in $\tau$ values across questions for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show larger values of $\tau$ than most other questions, as well as a larger absolute difference between people with a college degree and those without. Compared to the other questions explored here, change as adults plays a much larger role in explaining why people report the partisan identification and political ideology they do. And this is particularly true for respondents who do not have a college degree. The $\tau$ value for ideological identification among non-college educated respondents is almost four times that of college-educated respondents.

We note here that this meaningful difference in $\tau$ values across education groups and across questions would not be detectable using previous methods. For partisan identification and political ideology, both college-educated respondents and people with less than college degree would likely favor the AUM or LCAM over the SDM because they these questions both show evidence of some members of the population making intrapersonal change. In other words, using the tournament of models approach obscures the fact that the relative importance of intrapersonal change in explaining differences on these two items is different across these two groups, that intrapersonal change explains more systematic variance for ideological identification than partisan identification for people without a college degree, and that there appears to be more durable change on questions of affirmative action than on questions of government aid to black Americans.

We believe these patterns shed new light on the mechanisms underlying differences in attitude behavior across groups. Something about college attendance limits the ability of experiences later in life to drive changes in personal culture. While these results should not be interpreted as causal effects of attending college -- they are potentially confounded by age, social class, race, gender, and other factors that explain selection into higher education -- they open up a set of new questions and dynamics to explore.

@@

This paper set out intervene in the debate on whether people change their personal culture -- their attitudes, beliefs, values, practices, and dispositions -- as they move through their adult life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to proclaim victory for only one answer to this question, it seeks to refocus the debate to the theoretically more productive question of the relative importance of intrapersonal change for explaining differences in the personal culture. It proposes a new approach that quantifies the amount of observed systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change over the duration of a panel.

Applying our proposed measure, $\tau$, to all questions about personal culture from the panel datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions show evidence of people making durable, intrapersonal change over time, with some showing notably high intrapersonal change compared to interpersonal differences. For some questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

However, intrapersonal change is often substantially less pronounced than interpersonal differences, accounting for less than 10 percent of systematic variance on average across all datasets. On numerous questions, like those on civil liberties, abortion, generalized trust, and civic duty, the systematic variance attributable to intrapersonal change is essentially zero. For these questions, it seems there is not enough cultural change during adulthood to warrant attributing differences in personal culture mainly to experiences and social transitions; instead, the primary source of the observed differences appears to stem from experiences in childhood, adolescence, and early adulthood.

More than adjudicating past positions in this debate, the approach presented in this paper enables scholars to more clearly explore the relative contributions of change and stability in explaining variation in personal culture, as well as how these contributions might differ across populations. To exemplify this, we detailed how amount of cultural difference explained by intrapersonal change varies substantially by survey item but also by individuals' characteristics such as their education. Intrapersonal change plays almost no part in explaining sentiments regarding the government's economic role, but it is more defining for ideological and partisan identities. Intriguingly, it is more important in explaining cultural variation among non-college graduates compared to college graduates in the United States. While age and other factors related to college completion might confound this pattern, it suggests that college completion may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions less influential for college graduates relative to non-college graduates.

@@

Sociologists interested in understanding cultural differences have largely asked about the presence and significance of cultural change in adults [e.g., @kiley2020; @vaisey2021; @lersch2023]. Yet doing so has unnecessarily limited and might even set the wrong focus. In any population, some degree of cultural change in adults is inevitable. Likewise, no viable theoretical perspective would expect just one process -- stable differences or internal change -- to shape their personal culture. Our results reinforce this point, showing the relevance of both interpersonal differences and intrapersonal change in understanding the differences in personal culture among adults across all items analyzed.

This and other recent findings [@stewart2023; @quinn2023] suggest that it is theoretically more productive to ask about the relative importance of these two processes in understanding cultural variation. Drawing a unified conclusion from survey items measuring various cultural forms on different scales and across different time frames is challenging. However, the general pattern suggests that, for most items, intrapersonal change in adulthood is not the primary reason for cultural differences among people.

To clarify, our findings do not imply that adults remain static or that their changes are inconsequential. Even minor shifts in personal culture, such as a 2 percentage-point change in support for gay marriage, can have significant ramifications. While the majority may remain consistent in their views, understanding the underlying mechanisms of even such minor shifts remains a crucial task for the sociology of culture.

@@

This conclusion aligns with a range of recent causal inference work suggesting that selection effects, rather than treatment effects, predominantly account for personal cultural differences among individuals in varied social roles and positions [@campbell2016; @wodtke2018]. While some people clearly change as they transition into new roles or environments, this change often seems to be insufficient in magnitude and duration to explain what are often pronounced differences among people in diverse roles. [I'VE TRIED TO FIND A WAY TO GET BOTH IMPACT AND DURATION IN THERE; COULDN'T FIND A BETTER SOLUTION THOUGH--IDEAS? ALSO WE NEED TO BE CAREFUL WITH THIS SENTENCE, IT SEEMS WE ARE OVERSTATING HERE; LET'S FIND A WAY TO RELATIVIZE THE CLAIM HERE?]. These and our results suggest that when observing differences in personal culture across social roles, such as parenthood, education, or professional authority [@longest2013], or across occupations [@weeden2005], selection likely plays a large role in explaining these difference, though exceptions always exist.

Our analysis does not provide an answer as to why intrapersonal change seems to have limited impact on understanding cultural differences among adults. The situations that promote durable change in personal culture might simply be rare during adulthood. Alternatively, it is possible that adults do encounter opportunities, necessities, and incentives for change, but their ability or willingness to change decreases. All the more it is important to research when and how social situations can provoke durable change in adults.
